---
title: "Gsens"
author: "Jean-Baptiste Pingault & Frank Dudbridge"
date: "20 August 2020"
output: github_document
---
 

### Overview
#### [1: Introduction](#link1)
#### [2: One polygenic score case](#link2)
#### [3: Two polygenic scores](#link3)

 

# Introduction<a name="link1"></a>
The following script provides an example on how to run Gsens, a genetically informed sensitivity analysis. Please cite the following papers where the concept was first proposed and then implemented:  
1. Pingault, J.-B., O’Reilly, P. F., Schoeler, T., Ploubidis, G. B., Rijsdijk, F., & Dudbridge, F. (2018). Using genetic data to strengthen causal inference in observational research. Nature Reviews Genetics, 19(9), 566–580. https://doi.org/10.1038/s41576-018-0020-3  
2. (UPDATE) Pingault, J.-B., Rijsdijk, F., Schoeler, T., Choi, S. W., Kraphol, E., O’Reilly, P. F., & Dudbridge, F. (In Prep). Estimating the sensitivity of associations between risk factors and outcomes to shared genetic effects. BioRxiv.   

The following examples are based on the correlation matrix between polygenic scores and variables, which is available in the supplementary matrial of article 2. 

Set working directory and load gsens source code
```{r, results = "hide",message = FALSE  }
HOME=getwd()
setwd(HOME)
source('gsens.source.R')
``` 

Three functions are available:    
gsensY: sensivity based on one polygenic score for the outcome Y.  
gsensX: sensivity based on one polygenic score for the exposure X.  
gsensXY: sensivity based on two polygenic scores for X and Y.  

# One polygenic score case<a name="link2"></a>

##Observed scenario

The following example is based on maternal years of education (X), child GCSE (Y), and the best fitting polygenic score for years of education estimated in the child. 

```{r}
round(gsensY(rxy=0.3975,rgx = 0.2894,rgy = 0.3446,n=3785,h2=0.3446^2),3)
```

rxy, rgx, and rgy are the observed correlations between study variables (note that these should be partial correlations adjusting for sex, age, and principal components);   
n = sample size;  
h2 = is the variance explained in the outcome, here by the observed polygenic score (hence why h2 is rgy^2).    
The output provides:  

The standardized estimate of the relationship between X and Y, adjusted for G (i.e. the residual association after adjusting for the polygenic score). Note that this estimate should be the same as a regression of Y on X adjusting for the polygenic score in the dataset from which the correlations were obtained.  

The estimate of genetic confounding.  

The total effect which should add up to the observed initial association between X and Y.  

##Heritability scenario

The sensitivity analysis is implemented by providing the chosen heritability estimate, here h2 = 0.31 corresponding to the SNP-heritability of Y. 

```{r}
round(gsensY(rxy=0.3975,rgx = 0.2894,rgy = 0.3446,n=3785,h2=0.31),3)
```


As note in the manuscript, it is possible to fix the ratio k between rgy and rgx when a priori knowledge is available, for example the genetic relationship between the child and the mother. 

```{r}
 round(gsensY(rxy=0.3975,rgx = 0.5*sqrt(.31),rgy = sqrt(.31),n=3785,h2=0.31),3)
```

In this fixed solution, rgx and rgy and are set to their value under the heritability scenario, rather than the observed value from the polygenic score.

# Two polygenic scores <a name="link3"></a>

When a  polygenic score is available for each of X and Y, both can be modelled using gsensXY, for example,  for maternal years of education and BMI.

## Observed polygenic scores
```{r}
round(gsensXY(rxy=-0.0894,rg1x=0.2894,rg2y=0.2522,rg1y=-0.0268,rg2x=-0.0837,rg1g2=-0.1847,n=3663,h2.x=0.2894^2,h2.y=0.2522^2,print=T),3)
```
Similar to the one polygenic score  case the model is specified with paths between variables X and Y and polygenic scores for X (g1) and Y (g2). The option print = T enables the examination of model parameters. In addition to the negative variance, several parameters have unlikely values. For example, the cross path bg1y  flips from a negative to a positive value, which would imply that a higher polygenic score for education is linked to higher BMI. In such cases, constraints can be imposed on the the model in the following way.


```{r}

round(gsensXY(rxy=-0.0894,rg1x=0.2894,rg2y=0.2522,rg1y=-0.0268,rg2x=-0.0837,rg1g2=-0.1847,n=3663,h2.x=0.2894^2,h2.y=0.2522^2,print=T, constrain='lg1 < 1 \n lg2 < 1 \n  vg1 > 0  \n vg2 > 0 \n bg1y < 0 \n bg2x < 0 \n bxy < 0'),3)
```

Loadings for the latent part of the model are constrained to to be less than 1 and residual  variances to be positive. Cross paths and the residual association are constrained to be negative, ie not flip sign.  

Note that constraints need to be imposed cautiously on the model for two reasons:  

-To avoid nonsensical constraints, for example to keep the negative constraints on the cross paths imposed in this example in another situation where the polygenic scores are expected to be positively associated with both X and Y.

- Constraining parameters can decrease standard error even when the value of the parameter is not changed. That is, if lg1 is estimated to 1, constraining it to 1 will not change estimates but will prevent the model from estimating a standard error for lg1 and can reduce standard errors of other parameters including the estimates of interest.

Additional remarks regarding constraints:  
-Constraining vg1 > 0 did not prevent the warning regarding negative variance, which is estimated to -0.000. This can be solved by using vg1 > 1e-06. However, such negative variances of -0.000 can be left alone as constraining them does not change estimate values but can reduce standard errors. 
-All model parameters should be systematically checked after fitting a model, and before and after constraints are imposed. In addition to unplausible parameters, so-called 'heywood cases' should be checked. here, as the model is standardized, no parameter should be above 1 or below -1. For example, a heritability parameter above 1 would correspond to a heritability above 100%.



## Check with one polygenic score
Findings for the constrained vs the unconstrained model diverge. We can use gsensY to and just for the outcome-related polygenic score only, here BMI. Note that in theory, if the polygenic score for the outcome was perfect, adjusting for that polygenic score would be sufficient as it would capture  shared genetic influences between X and Y. Findings from adjusting for the BMI polygenic score only converge with the constrained version of the two polygenic score appraoch.

```{r}
round(gsensY(rxy=-0.0894,rgx = -0.0837,rgy = 0.2522,n=3663,h2=0.2522^2),3)
```

Note here that the vg is -0.000 here too, which does not affect the model so the warning can be safely ignored.
 
##Heritability scenario


```{r}
round(gsensXY(rxy=-0.0894,rg1x=0.2894,rg2y=0.2522,rg1y=-0.0268,rg2x=-0.0837,rg1g2=-0.1847,n=3785,h2.x=0.25*0.31,h2.y=0.186,
        constrain='lg1 < 1 \n lg2 < 1 \n  vg1 > 0  \n vg2 > 0 \n bg1y < 0 \n bg2x < 0 \n bxy < 0'),3)
```

Heritability scenarios of interest can be modelled we two polygenic scores by replacing h2.x and h2.y by the chosen values, here SNP-heritability estimates. Not that in h2.x=0.25x0.31, the factor 0.25 corresponds to the genetic relatedness between child and mother (i.e. sqrt(h2.x) is computed in the model leading to the value of the path equal to 0.5xsqrt(0.31).) 


 
 